Performance portability of an intermediate-complexity atmospheric research model in coarray Fortran

### 2017-12-11 body.tex
Final submisssion.

\subsection{Objectives}
This evaluates alternatives for compiling, linking, and executing one
\gls{mini-app} \gls{caf} source code using the following technologies:
\begin{itemize}
  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{openshmem2016} communication layers,
  \item \gls{gcc} and Cray compilers, and
  \item Multi- and many-core processors.
\end{itemize}
We also tested the Intel Fortran compiler versions 16 and 17. The resulting executable
program crashes on Cray platforms and hangs at launch on when run on more than 16 cores on the SGI platform.

We show performance for two one-sided data access patterns:
\begin{enumerate}
  \item Gets: an image retrieves data from memory managed by another
        image without the providing image's involvement, and
  \item Puts: an image stores data in memory managed by
        another image without the receiving image's involvement.
\end{enumerate}
Fortran semantics necessitates that gets block.  Puts are non-blocking.

%Our results will inform our future decisions around the choice of compiler, communication
%layer, platform, and data access patterns.

@@

The \gls{ncar} system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, each containing two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
Cheyenne uses a Mellanox EDR Infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
We compiled coarray-\gls{icar} on the \gls{nersc} systems using the Cray Fortran compiler version 8.6.0.  We compiled
at \gls{ncar} using the \gls{gcc} version 6.3 Fortran front end, which uses the
OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} to support \gls{caf}.  We tested two OpenCoarrays parallel runtime libraries:
\begin{enumerate}
  \item The default \gls{mpi} library using the SGI MPT \gls{mpi},
  \item The recently released OpenSHMEM library.
\end{enumerate}
We distributed \gls{caf} images as follows:
\begin{itemize}
  \item 68 images per Xeon Phi node on Cori,
  \item 24 images per Xeon node on Edison, and
  \item 36 images per node on Cheyenne,
\end{itemize}
corresponding to one image per physical core in each case.

@@

Figure~\ref{fig3-4} (top) depicts results across multiple compilers, machines, and architectures.
In this case, only ``puts'' are tested, and only for the $2000 \times 2000\times 20$ domain.
The GNU compiler was used on Cheyenne (Xeons). The Cray compiler was used on Edison (Xeons) and Cori (KNL).
To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
These results show that the Cray compiler+system scales better than the gfortran+SGI system,
with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
This also shows that the KNL system scales well out to large core counts (60\% ideal with \num{19200} cores),
but that the total runtimes are significantly slower than the equivalent runtimes on Xeons (top right).
The KNL performance might be improved in the future by implementing OpenMP threaded parallelism within a node.

>>>>

\subsection{Objectives}
This paper evaluates alternatives for compiling, linking, and executing one
\gls{mini-app} \gls{caf} source code using the following technologies:
\begin{itemize}
  \item \gls{mpi}~\cite{mpiforum2016mpi} and OpenSHMEM~\cite{openshmem2016} communication layers,
  \item \gls{gcc} and Cray compilers, and
  \item Multi- and many-core processors.
\end{itemize}
We also tested the Intel Fortran compiler versions 16 and 17. The resulting executable
program crashes on Cray platforms and hangs at launch on when run on more than 16 cores on the SGI platform.

We show performance for two one-sided data access patterns:
\begin{enumerate}
  \item Gets: an image retrieves data from memory managed by another
        image without the providing image's involvement, and
  \item Puts: an image stores data in memory managed by
        another image without the receiving image's involvement.
\end{enumerate}
We also explore the performance and scalability of multi- versus many-core processors.

@@

The \gls{ncar} system is ``Cheyenne,'' an SGI ICE XA Cluster with \num{4032} compute nodes, each containing two sockets of 18-core Intel Xeon Processor E5-2697 v4 (``Broadwell''), running at \num{2.3}~\si{\giga\hertz}.
Cheyenne uses a Mellanox EDR Infiniband interconnect with a partial 9D Enhanced Hypercube single-plane topology.
We compiled coarray-\gls{icar} on the \gls{nersc} systems using the Cray Fortran compiler version 8.6.0.  We compiled
at \gls{ncar} using the \gls{gcc} version 6.3 Fortran front end, which uses the
OpenCoarrays \gls{abi}~\cite{fanfarillo2014opencoarrays} to support \gls{caf}.  We tested two parallel runtime libraries that implement the OpenCoarrays \gls{abi}:
\begin{enumerate}
  \item The default \gls{mpi} library using the SGI MPT \gls{mpi},
  \item The recently released OpenSHMEM library.
\end{enumerate}
We distributed \gls{caf} images as follows:
\begin{itemize}
  \item 68 images per Xeon Phi node on Cori,
  \item 24 images per Xeon node on Edison, and
  \item 36 images per node on Cheyenne,
\end{itemize}
corresponding to one image per physical core in each case.

@@

Figure~\ref{fig3-4} (top) depicts results across multiple compilers, machines, and architectures.
In this case, only ``puts'' are tested, and only for the $2000 \times 2000\times 20$ domain.
The GNU compiler was used on Cheyenne (Xeons). The Cray compiler was used on Edison (Xeons) and Cori (KNL).
To aid interpretation, the scaling results are reported as a fraction of the ideal scaling for each machine in the top left.
These results show that the Cray compiler+system scales better than the gfortran+SGI system,
with 75\% efficiency at >10k cores, while on Cheyenne, only 55\% of ideal was achieved.
This also shows that the KNL system scales well out to large core counts (60\% of the ideal with \num{19200} cores),
but that the total runtimes are significantly slower than the equivalent runtimes on Xeons (top right).
The KNL performance might be improved in the future by implementing OpenMP threaded parallelism within a node.

### 2017-12-11 main.tex
Final submisssion.

\begin{abstract}
We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
several parallel, numerical algorithms known to dominate the execution of \gls{icar}~\cite{gutmann2016intermediate} model,
a package developed at the \gls{ncar}.
The \gls{mini-app} uses standard Fortran 2008, including several Fortran 2008 implementations of the collective
subroutines defined in the Committee Draft of the upcoming Fortran 2015 standard.  The ability of \gls{caf} to run atop various
communication layers and the increasing \gls{caf} compiler availability facilitated evaluating several compilers,
runtime libraries and hardware platforms.  Results are presented for the GNU and Cray compilers, each of which offers
different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
alternatives.  We studied performance on multi- and many-core processors in distributed memory.  The
results show promising scaling behavior across a range of hardware, compiler, and runtime choices
on platforms ranging up to $\sim$100,000 cores.
\end{abstract}

>>>>

\begin{abstract}
We examine the scalability and performance of an open-source, \gls{caf} \gls{mini-app} that solves
several parallel, numerical algorithms known to dominate the execution of \gls{icar}~\cite{gutmann2016intermediate} model,
a package developed at the \gls{ncar}.
The \gls{mini-app} uses standard Fortran 2008, including one Fortran 2008 implementation of a collective
subroutine defined in the Committee Draft of the upcoming Fortran 2018 standard.  The ability of \gls{caf} to run atop various
communication layers and the increasing \gls{caf} compiler availability facilitated evaluating several compilers,
runtime libraries and hardware platforms.  Results are presented for the GNU and Cray compilers, each of which offers
different parallel runtime libraries employing one or more communication layers, including \gls{mpi}, OpenSHMEM, and proprietary
alternatives.  We studied performance on multi- and many-core processors in distributed memory.  The
results show promising scaling behavior across a range of hardware, compiler, and runtime choices
on platforms ranging up to $\sim$100,000 cores.
\end{abstract}
